Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2016-03-17 rmarkdown/paper.Rnw
added mappability and gc content scores part to methods and added the corresponding bibliography entries

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of \Sexpr{bin_size} non-overlapping
window after extending reads by \Sexpr{frag_len} to their $3\prime$
end directions. ChIP tag counts in ChIP-exo data were linearly related
to ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

We next evaluated the ``peak pair'' assumption from Rhee and Pugh,
2011 \cite{exo1}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site. Wang et al., 2014
\cite{mace}, Madrigal 2015 \cite{cexor} and Bardet et al., 2013
\cite{peakzilla} proposed method rely in this assumption. In order to
evaluate this assumption, we reviewed the proportion of reads in the
forward strand in candidate regions (i.e. regions with at least one
binding site) in $\sig$ ChIP-exo data. We found that strands of reads
were much less balanced in ChIP-exo data than in ChIP-Seq data in
these regions with potential binding sites (Fig. \ref{fig:comp}B) and
this indicates that the peak pair assumption might not hold in real
ChIP-exo data.

We evaluated ChIP-exo data for CTCF factor from human genome
\cite{exo1} to investigate issues specific to eukaryotic genomes for
binding sites identification. Figures \ref{fig:comp}C and
\ref{fig:comp}D display the bin-level average read counts against
mappability and GC content. Each data point is obtained by averaging
the read counts across bins with the same mappability of GC
content. In Figure \ref{fig:comp}C it is shown that the ChIP-exo tag
counts linearly increases with the mappability score and in Figure
\ref{fig:comp}D it is shown that for GC - content below 0.6, the mean
ChIP tag count increases and for GC - content greater than 0.6 it
shows a decreasing trend.These results indicate that binding site
identification in ChIP-exo sample might also benefit from the use of
methods that take into account of apparent sequence biases such as
mappability and GC content.

@@

\begin{align*}
  \text{ARC} &= \dfrac{\text{Nr. of reads in the region}}{\text{Width of the region}} \\
  \text{URCR} &= \dfrac{\text{Nr. of reads in the region mapped to
      exactly one position}}{\text{Nr. of reads in the region}}
\end{align*}

@@

@@

\subsection*{ChIP-Seq quality control metrics and Strand
  Cross-Correlation.}

The statistics in Table \ref{tab:qc} and the SCC curves from Figure
\ref{fig:scc_exo} where calculated with the \textbf{ChIPUtils} package
version 0.99.0, available in
\url{https://github.com/welch16/ChIPUtils}.

@@

\begin{figure}[h!]
\centering
\includegraphics[width = .8\textwidth]{/p/keles/ChIPexo/volume3/ChIPexo/figs/for_paper/EColi_strand_cross_corr.pdf}
\caption{SCC curves for $\sig$ samples. The ``phantom peak'' and the
  summit that corresponds to the read and fragment length respectively
  are confounded due to the exonuclease digestion.}
  \label{fig:scc_exo}
\end{figure}

>>>>

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of \Sexpr{bin_size} non-overlapping
window after extending reads by \Sexpr{frag_len} to their $3\prime$
end directions. ChIP tag counts in ChIP-exo data were linearly related
to ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Upper part of Figure \ref{fig:comp}A). This implies that
signals for potential binding sites are well reproducible between
ChIP-exo and ChIP-Seq data. On the other hand, there was clear
difference in the background distribution between them (lower part of
Figure \ref{fig:comp}A). Specifically, in ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

We next evaluated the ``peak pair'' assumption from Rhee and Pugh,
2011 \cite{exo1}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site. Note that currently
available ChIP-exo data analysis methods, such as Wang et al., 2014
\cite{mace}, Madrigal 2015 \cite{cexor} and Bardet et al., 2013
\cite{peakzilla} rely on this assumption. In order to evaluate this
assumption, we reviewed the proportion of reads in the forward strand
in ChIP-exo peaks such as at least one binding site is predicted in
both ChIP-exo and ChIP-Seq data. We found that strands of reads were
much less balanced in ChIP-exo data than in ChIP-Seq data in these
regions with potential binding sites (Fig. \ref{fig:comp}B) and this
indicates that the peak pair assumption might not hold in real
ChIP-exo data.

We evaluated ChIP-exo data for CTCF factor from human genome
\cite{exo1} to investigate issues specific to eukaryotic genomes for
binding sites identification. Figures \ref{fig:comp}C and
\ref{fig:comp}D display the bin-level average read counts against
mappability and GC content. Each data point is obtained by averaging
the read counts across bins with the same mappability of GC
content. In Figure \ref{fig:comp}C it is shown that the ChIP-exo tag
counts linearly increases with the mappability score and in Figure
\ref{fig:comp}D it is shown that for GC - content below 0.6, the mean
ChIP tag count increases and for GC - content greater than 0.6 it
shows a decreasing trend.

These results indicate that binding site identification in ChIP-exo
sample might also benefit from the use of methods that take into
account of apparent sequence biases such as mappability and GC
content.

@@

\begin{align*}
  \text{ARC} &= \dfrac{\text{Nr. of reads in the region}}{\text{Width of the region}}, \\
  \text{URCR} &= \dfrac{\text{Nr. of reads in the region mapped to
      exactly one position}}{\text{Nr. of reads in the region}}.
\end{align*}

@@

The statistics in Table \ref{tab:qc} and the SCC curves from Figure
\ref{fig:scc_exo} were calculated with the \textbf{ChIPUtils} package
version 0.99.0, available in
\url{https://github.com/welch16/ChIPUtils}.

@@

\subsection*{Mappability and GC content scores.}

To define the mappability score we follow the definition from Rozowsky
et al., 2009 \cite{peakseq}:

\begin{align}
  m_i &= \sum_{k = i - L +1}^{i + L - 1} \frac{\delta_k }{2L - 1}.
\end{align}

where $\delta_i$ is the indicator if the base at coordinate $i$ can be
mapped uniquely by a 32 bp sequence at position $i$, and $L$ is the
expected fragment length. GC - content score is defined analogously,
where $\delta_i$ represents the ocurrence of a G or C at the $i$-th
position in the genome.

The mappability and GC - content scores for a bin are defined as the
average of the scores across the nucleotides in the bin.

@@

\begin{figure}[h!]
\centering
\includegraphics[width = .8\textwidth]{/p/keles/ChIPexo/volume3/ChIPexo/figs/for_paper/EColi_strand_cross_corr.pdf}
\caption{\textbf{SCC curves for $\sig$ samples.} The ``phantom peak''
  and the summit that corresponds to the read and fragment length
  respectively are confounded due to the exonuclease digestion.}
  \label{fig:scc_exo}
\end{figure}

### 2016-03-17 rmarkdown/paper.tex
added mappability and gc content scores part to methods and added the corresponding bibliography entries

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their $3\prime$
end directions. ChIP tag counts in ChIP-exo data were linearly related
to ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
potential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

We next evaluated the ``peak pair'' assumption from Rhee and Pugh,
2011 \cite{exo1}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site. Wang et al., 2014
\cite{mace}, Madrigal 2015 \cite{cexor} and Bardet et al., 2013
\cite{peakzilla} proposed method rely in this assumption. In order to
evaluate this assumption, we reviewed the proportion of reads in the
forward strand in candidate regions (i.e. regions with at least one
binding site) in $\sig$ ChIP-exo data. We found that strands of reads
were much less balanced in ChIP-exo data than in ChIP-Seq data in
these regions with potential binding sites (Fig. \ref{fig:comp}B) and
this indicates that the peak pair assumption might not hold in real
ChIP-exo data.

We evaluated ChIP-exo data for CTCF factor from human genome
\cite{exo1} to investigate issues specific to eukaryotic genomes for
binding sites identification. Figures \ref{fig:comp}C and
\ref{fig:comp}D display the bin-level average read counts against
mappability and GC content. Each data point is obtained by averaging
the read counts across bins with the same mappability of GC
content. In Figure \ref{fig:comp}C it is shown that the ChIP-exo tag
counts linearly increases with the mappability score and in Figure
\ref{fig:comp}D it is shown that for GC - content below 0.6, the mean
ChIP tag count increases and for GC - content greater than 0.6 it
shows a decreasing trend.These results indicate that binding site
identification in ChIP-exo sample might also benefit from the use of
methods that take into account of apparent sequence biases such as
mappability and GC content.

>>>>

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. In order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts across the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their $3\prime$
end directions. ChIP tag counts in ChIP-exo data were linearly related
to ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Upper part of Figure \ref{fig:comp}A). This implies that
signals for potential binding sites are well reproducible between
ChIP-exo and ChIP-Seq data. On the other hand, there was clear
difference in the background distribution between them (lower part of
Figure \ref{fig:comp}A). Specifically, in ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that for ChIP-exo data a much smaller portion of the
genome is expected to be background.

We next evaluated the ``peak pair'' assumption from Rhee and Pugh,
2011 \cite{exo1}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site. Note that currently
available ChIP-exo data analysis methods, such as Wang et al., 2014
\cite{mace}, Madrigal 2015 \cite{cexor} and Bardet et al., 2013
\cite{peakzilla} rely on this assumption. In order to evaluate this
assumption, we reviewed the proportion of reads in the forward strand
in ChIP-exo peaks such as at least one binding site is predicted in
both ChIP-exo and ChIP-Seq data. We found that strands of reads were
much less balanced in ChIP-exo data than in ChIP-Seq data in these
regions with potential binding sites (Fig. \ref{fig:comp}B) and this
indicates that the peak pair assumption might not hold in real
ChIP-exo data.

We evaluated ChIP-exo data for CTCF factor from human genome
\cite{exo1} to investigate issues specific to eukaryotic genomes for
binding sites identification. Figures \ref{fig:comp}C and
\ref{fig:comp}D display the bin-level average read counts against
mappability and GC content. Each data point is obtained by averaging
the read counts across bins with the same mappability of GC
content. In Figure \ref{fig:comp}C it is shown that the ChIP-exo tag
counts linearly increases with the mappability score and in Figure
\ref{fig:comp}D it is shown that for GC - content below 0.6, the mean
ChIP tag count increases and for GC - content greater than 0.6 it
shows a decreasing trend.

These results indicate that binding site identification in ChIP-exo
sample might also benefit from the use of methods that take into
account of apparent sequence biases such as mappability and GC
content.
